# Building a DAG

Blindly applying machine learning to geology and geometallurgy may work. But as soon as we try to apply any form of causal thinking to the resulting statistical models bad things can happen. The Simpson's paradox and Berkson's paradox illustrate the dangers.

Machine learning can replicate the joint conditional probabilities of the training data closely even if the statistical model is miss-specified (missing covariates). No matter what we do in trying to build a good statistical model by using techniques such as data splitting for controlling model over-fitting or under-fitting--miss-specified statistical models are still miss-specified statistical models and they are still wrong and can lead to wrong predictions, especially when predicting beyond the domain (range) of the statistical model.

```{r}
#| output: false
library(simDAG)
library(tidyverse)
library(dplyr)
library(dagitty)
library(tidyr)
library(stringr)
library(ggdag)
library(ggplot2)

source("./r/plot_dag.R")
source("./r/rtruncnorm.R")
```

## Geology example

In mineral resource estimation, the use of machine learning comes in handy when dealing with missing values or with dealing with highly under-sampled data such as the bulk density of rock samples. For bulk density a common approach is to simply use assays to predict density. Assays have molar mass which is directly correlated with bulk density ($assays \rightarrow bulk\ density$) of a rock sample.

To simplify the visual presentation of the DAG, the DAG nodes will be abbreviated.

DAG node abbreviations:

- X = assays
- Y = rock bulk density
- L = lithology
- M = mineral assemblage
- R1 = porosity
- R2 = moisture
- R3 = texture
- R4 = hardness
- R5 = rock competence
- G1 = weathering
- G2 = alteration
- G3 = mineralisation
- G4 = oxidation

Bulk density of a rock sample can be estimated mathematically by the additive volume of its individual mineral phases. This leads to $mineral\ assemblage \rightarrow bulk\ density$. The mineral assemblage determines the geochemical composition of the rock ($mineral\ assemblage \rightarrow bulk\ assays$).

The initial DAG is:

$$bulk\ density\ (Y) \leftarrow mineral\ assemblage\ (M) \rightarrow assays\ (X) \rightarrow bulk\ density\ (Y)$$

Bulk density of the rock sample is the outcome (Y), assays is the exposure (X), and the mineral assemblage is the moderator (M).

What about other geological factors that we might have observations or measurements for that may affect rock bulk density? Typical geological factors may include: lithology (discrete), alteration(discrete), oxidation (discrete), weathering (discrete), moisture content (continuous), porosity (continuous), rock hardness (continuous), rock competence (measured through RQD), texture, and spatial coordinates X, Y, and Z (continuous).

Of the available geological factors, porosity and moisture content are direct physical determinants of density (at fixed solid mineral composition). We can add the edges $porosity\ (R1) \rightarrow Y$ and $moisture\ (R2) \rightarrow Y$ to the DAG.

What geological factors control porosity and moisture? Porosity is controlled by the observed rock features texture, lithology (via texture) and geological processes weathering, alteration (which also affects texture), mineralisation, and oxidation. Moisture content is controlled by weathering. We now add the edges:

- $weathering\ (G1) \rightarrow porosity\ (R1)$
- $alteration\ (G2) \rightarrow porosity\ (R1)$
- $oxidation\ (G4) \rightarrow porosity\ (R1)$
- $texture\ (R3) \rightarrow porosity\ (R1)$
- $lithology\ (L) \rightarrow texture\ (R3) \rightarrow porosity\ (R1)$
- $weathering\ (G1) \rightarrow moisture\ content (R2)$.

Lithology also directly affects mineral assemblage and texture. We need to also add the edges:

- $lithology (L) \rightarrow mineral\ assemblage\ (M)$
- $lithology (L) \rightarrow texture\ (R3)$
- $weathering (G1) \rightarrow mineral\ assemblage\ (M)$
- $alteration (G2) \rightarrow mineral\ assemblage\ (M)$
- $mineralisation (G3) \rightarrow mineral\ assemblage\ (M)$
- $oxidation (G4) \rightarrow mineral\ assemblage\ (M)$.

Alteration also directly affects rock texture and geochemistry via changing mineral assemblage. we need to add the edges:

- $alteration (G2) \rightarrow teture (R3)$
- $alteration (G2) \rightarrow mineral\ assemblage\ (M)$.

The geological processes oxidation and weathering are strongly linked with oxidation affecting weathering or weathering affecting oxidation (depends on your conceptual choice).

The geological features hardness ($R4$) and rock competence ($R5$) are more symptoms than primary causes of rock bulk density. Lithology, weathering, and alteration affect rock hardness while porosity, texture and weathering all affect rock competence.

Finally, we can capture the spatial distribution of density via X, Y, and Z coordinates. These variables are freely available via geological interpretation. Both X and Y coordinates drive the rock properties while Z could be considered a direct effect as depth determines the overall mineral assemblage via lithology, alteration, weathering and oxidation states and the range limits for porosity and moisture content.

Let's not forget about geological domains. For mineral resource estimation, geological domains are a consequence of geological three-dimensional interpretation of the data. Geological domains are considered selection variables and typically factors in lithology, mineralisation, alteration, oxidation, and grade. How the geological domains are defined determines what edges (connections) are added to the DAG.

The resultant geology DAG for predicting rock bulk density from rock geochemistry (assays) is presented in @fig-dag-geo1. This graph can be simplified to @fig-dag-geo2 where $X$ is the rock geochemistry, $Y$ is rock bulk density, $L$ is lithology, $G$ is a latent variable representing geological processes (weathering, alteration, mineralisation, and oxidation), $R$ is a latent variable representing rock properties (porosity, moisture, rock texture, rock hardness, and rock competence), and $S$ is domain.

```{r}
#| label: fig-dag-geo1
#| fig-cap: "Assays -> Rock Bulk Density DAG"
#| fig.width: 6
#| fig.height: 4
dagify(
  R1 ~ G1 + G2 + G3 + R3,
  R2 ~ G1,
  R3 ~ L + G2,
  R4 ~ L + G1 + G2,
  R5 ~ G1 + R1 + R3,
  G4 ~ G1,
  M ~ L + G1 + G2 + G3 + G4,
  X ~ M,
  Y ~ X + M + R1 + R2
) |> plot_dag(layout="stress")
```

```{r}
#| label: fig-dag-geo2
#| fig-cap: "Assays -> Rock Bulk Density DAG"
#| fig.width: 6
#| fig.height: 4
dagify(
  M ~ L + G,
  R ~ L + G,
  X ~ M,
  Y ~ X + M + R,
  S ~ X + G + L,
  labels = c(
    M = "Mineral Assemblage",
    L = "Lithology",
    G = "Geological Processes",
    R = "Rock Properties",
    X = "Assays",
    Y = "Bulk Density",
    S = "Domain"
  )
) |> plot_dag(layout="stress") +
  geom_dag_label_repel(
    aes(label=label),
    segment.alpha = 1,
    size = 3,
    show.legend = FALSE,
    box.padding = grid::unit(2, "lines"),
    label.padding = grid::unit(0.2, "lines")
  )
```

For most data sets used for training a predictive rock bulk density model, the mediators mineral assemblage and rock properties are either completely missing (most likely) or partly available for a subset of the mineral resource project area--spatial coverage is usually poor. To over come this we resort to using secondary variables (logged lithology and geological processes) that can stand in for mineral assemblage and rock properties.

If we don't have these secondary variables and all we have is $X$ and $Y$, then we need to insert one or more latent variables that represent mineral assemblage, lithology, geological processes and rock properties. This might be domain or a latent variable that represents mineral assemblage, lithology, rock properties, and geological processes. The DAG can then collapse to $S \rightarrow X \rightarrow Y \leftarrow S$. $S$ needs to be a structural domain that is a latent cause of geology and rock properties and not a selection/estimation domain which determines how the data is carved up for modelling and spatial estimation.

How domain is treated becomes a very important consideration when trying to accurately predict density. The domain $S$ constructed directly from the data ($S = f(M, R, G, C)$) using typical geological interpretation methods means that the DAG is $S \leftarrow X \rightarrow Y \rightarrow S$. This is a selection domain. If we train a classifier to predict membership from the data ($S^* = \text{classifier}(X, M, R, G, C)$), then the DAG becomes $S^* \leftarrow X \rightarrow Y \rightarrow S*$. This is a latent domain. These two "domains" are not the same ($S \neq S^*$). An example of $S^*$ is using latent variable modelling or Gaussian mixture modelling to infer the latent domain classes (labels).

Not all is lost. If we have a domain ($S$) which is a selection variable and we want to use Bayesian Networks to model the data, we can employ the mixture‑of‑Gaussians Bayesian Network trick to reverse the links $X,Y \rightarrow S$ to $S \rightarrow X,Y$. The trick is similar to the examples for modelling non-linear relationships but instead a latent Gaussian variable and a selection variable are added to the DAG.

## Geometallurgy example

From metallurgy testwork data it is common to develop grade-recovery curves. For instance, for a gold deposit, recovery is estimated from gold head grade using a non-linear formula such as $Y = K_1(1 - exp(-K_2*X)) + K_3log(X) + K_4$ or $Y = \dfrac{K_1}{X} + K_2$. But this assumes that the we are interested in predicting $Y$ given $X$ ($P(Y|X)$). It implies that there is no causal relationship between $X$ and $Y$, only a correlation.

This approach is sufficient if we are adding predictions to a block model to be used for Resource or Reserve estimation.

It fails if we are interested in using grade as a parameter to control recovery, that is, to blend different material to give a target grade as is common practice in ore control. For this we should be using $P(Y|do(X))$ (do calculus). The $P(Y|do(X))$ means that we are interested in predicting $Y$ given that we set a specific value for $X$. The distribution $P(Y|X)$ is not the same as $P(Y|do(X))$. This now implies a causal relationship between $X$ and $Y$.

Determining true causal relationships in geometallurgy is both expensive and time consuming and requires a researchers mindset.

The simDAG R package is used to simulate a geometallurgy data set with the DAG $X \leftarrow U \rightarrow Y$ where $X$ is grade, $Y$ is metal recovery, and $U$ is the missing geological and metallurgical confounders. There is no causal relationship between $X$ and $Y$. The true (simulated) cause $U$ is missing and is a representation of potentially multiple interacting factors. Using a low number of samples replicates the small size typical of geometallurgy data sets.

```{r}
offsets_x <- c("1"=1.52,"2"=0.64,"3"=10.66,"4"=0.64,
               "5"=0.82,"6"=22.9,"7"=1.13)
sigma_x <- c("1"=1.15,"2"=0.39,"3"=4.52,"4"=0.01,
             "5"=0.63,"6"=12.57,"7"=0.81)
node_x <- function(data, parents, betas=NULL) {
  m <- offsets_x[data$U]
  z <- rtruncnorm(nrow(data), mean=m, sd=sigma_x[data$U], a=0.001)
  return(z)
}

offsets_y <- c("1"=82.65,"2"=63.31,"3"=82.55,"4"=35.08,
               "5"=48.64,"6"=91.58,"7"=73.04)
slope_y <- c("1"=0.78,"2"=0.16269347,"3"=0.02,"4"=0.01,
             "5"=4.44,"6"=-0.14,"7"=0.42)
sigma_y <- c("1"=3.24,"2"=2.6,"3"=1.99,"4"=0.01,
             "5"=3.24,"6"=1.9,"7"=2.88)
node_y <- function(data, parents, betas=NULL) {
  m <- offsets_y[data$U] + slope_y[data$U]*data$X
  z <- rtruncnorm(nrow(data), mean=m, sd=sigma_y[data$U], a=0.001)
  return(z)
}

dag_geomet <- empty_dag() +
  node("U", type="rcategorical", probs=c(0.64,0.07,0.02,0.01,0.02,0.01,0.23),
       output="factor", labels=1:7) +
  node("Z", type="rnorm", mean=0, sd=1) +
  node("X", type=node_x, parents="U") +
  node("Y", type=node_y, parents="U")

# Simulate the Naive Measured Dataset
set.seed(1700)
df_geomet <- sim_from_dag(dag_geomet, n_sim=100) |> 
  as.data.frame() |>
  mutate(Source = "Measured") |>
  select(X, Y)
```

The synthesized recovery on grade is shown in @fig-geomet1. The plot shows the measured scatter plot coloured by point density along with a non-linear equation that approximates the data (solid black line) and two boundary limits (black dashed lines). Note that even though the grade $X$ has no relationship with $Y$, $X$ is being used to predict $Y$. The correlation between $X$ and $Y$ is 0.33 while $log(X)$ and $Y$ is 0.42.

Note that this example is fabricated but the relationship between grade and metal recovery can be strongly influenced by the interaction of recovered mass, recovered metal, and grade of the concentrate. But the grade of the concentrate is a different variable from grade of the rock being fed into the process plant.

```{r}
#| label: fig-geomet1
#| fig-cap: "Assays -> Rock Bulk Density DAG"
#| fig.width: 5
#| fig.height: 5

v <- densCols(df_geomet, colramp=colorRampPalette(c("black", "white")))
v <- col2rgb(v)[1,] + 1L
cols <- viridis::viridis(256, alpha=1)
cols <- cols[v]

plot(df_geomet$X, df_geomet$Y, ylim=c(0,100), col=cols, pch=16, xlim=c(0.01,15),
     main="Synthesized Geomet data", xlab="Grade (g/t)", ylab="Metal Recovery (%)")

xx <- c(0.001, seq(0.01, 25, by=0.01))
yy <- 91*(1-exp(-20*xx)) + 0.9*log(xx)
lines(xx, yy, lty="dashed", col="black")
yy <- 84*(1-exp(-5*xx)) + 1*log(xx)
lines(xx, yy, lty="solid", col="black", lwd=2)
yy <- 75*(1-exp(-0.6*xx)) + 2*log(xx)
lines(xx, yy, lty="dashed", col="black")
```

## Crafting a DAG from a list of mineral formula

The general structure of the DAG is $mineral\ group \rightarrow mineral \rightarrow element$.

To assist construction of the DAG in DOT notation two helper functions `classify_mineral_group` (classify mineral into mineral group based on anion formula) and `create_mineral_dag` (build the DAG) are defined.

```{r}
classify_mineral_group <- function(formula) {
  case_when(
    # Native Minerals: single element symbol (e.g., Au, Cu, C, Ag)
    str_detect(formula, "^[A-Z][a-z]?[0-9]*$") ~ "Native Minerals",
    
    # Silicates: contains Silicon (Si)
    str_detect(formula, "Si") ~ "Silicates",
    
    # Carbonates: contains CO3
    str_detect(formula, "CO3") ~ "Carbonates",
    
    # Phosphates: contains PO4 or P with O
    str_detect(formula, "PO4|P") ~ "Phosphates",
    
    # Sulphates: contains SO4 group
    str_detect(formula, "SO4") ~ "Sulphates",
    
    # Sulphides: contains S (without SO4 group)
    str_detect(formula, "S") ~ "Sulphides",
    
    # Halides: contains Halogens (F, Cl, Br, I) without Oxygen
    str_detect(formula, "F|Cl|Br|I") & !str_detect(formula, "O") ~ "Halides",
    
    # Oxides & Hydroxides: contains Oxygen (O/OH) not captured above
    str_detect(formula, "O|OH") ~ "Oxides",
    
    TRUE ~ "Other"
  )
}

create_mineral_dag <- function(data) {
  pattern <- "H2O|CO3|CO2|OH|[A-Z][a-z]?"
  
  # Clean names & classify into Mineral Groups
  processed_df <- data %>%
    mutate(
      mineral_clean = str_replace_all(mineral, "[^a-zA-Z0-9_]", "_"),
      mineral_group = classify_mineral_group(formula)
    )
  
  # Layer 1 Edges (Mineral Group -> Mineral)
  group_edges <- processed_df %>%
    distinct(mineral_group, mineral_clean) %>%
    mutate(
      group_clean = str_replace_all(mineral_group, "[^a-zA-Z0-9_]", "_"),
      edge_str = paste0("  ", group_clean, " -> ", mineral_clean)
    )
  
  # Layer 2 Edges (Mineral -> Element / Polyatomic Node)
  element_edges <- processed_df %>%
    mutate(nodes = str_extract_all(formula, pattern)) %>%
    unnest(cols = c(nodes)) %>%
    distinct(mineral_clean, nodes) %>%
    mutate(edge_str = paste0("  ", mineral_clean, " -> ", nodes))
  
  # Combine Edges into DOT Notation
  all_edges <- c(group_edges$edge_str, element_edges$edge_str)
  dot_notation <- paste0("dag {\n", paste(all_edges, collapse = "\n"), "\n}")
  
  # Parse into dagitty object
  dag_obj <- dagitty(dot_notation)
  
  return(list(
    dot = dot_notation,
    data = processed_df,
    dag = dag_obj
  ))
}
```

Now for an example DAG. Note that visualizing the graph can get messy very quickly as the size of the graph grows. For this small example the graph is close to being unreadable. Any bigger and the labels will start to over-plot each other and the edges (links) will be obscured or become too short to be plotted.

```{r}
# Test dataset across all 8 mineral groups
df <- data.frame(
  mineral = c("Quartz", "Hematite", "Calcite", "Galena", 
              "Gypsum", "Halite", "Apatite", "Gold", "Malachite"),
  formula = c("SiO2", "Fe2O3", "CaCO3", "PbS", 
              "CaSO4·2H2O", "NaCl", "Ca5(PO4)3F", "Au", "Cu2(OH)2CO3")
)

result <- create_mineral_dag(df)

# Print generated DOT string
cat("=== 3-Tier DOT Notation Output ===\n")
cat(result$dot)
cat("\n==================================\n\n")
```

The resulting DAG is shown in @fig-min-dag.

```{r}
#| label: fig-min-dag
#| fig-cap: "Example DAG from mineral formula"
#| fig.width: 8
#| fig.height: 6

# --- Plot ---
tidy_dagitty(result$dag, layout="stress") |>
  ggplot(aes(x=x, y=y, xend=xend, yend=yend)) +
  geom_dag_edges_link(aes(start_cap=ggraph::circle(2,"mm"),
                          end_cap=ggraph::circle(2,"mm"))) +
  geom_dag_label(aes(label=name), 
                       color="black", 
                       fill="lightblue",
                       size=3) +
  theme_dag() +
  labs(
    title = "3-Tier Mineral Relationship DAG",
    subtitle = "Mineral Group  →  Mineral Species  →  Chemical Node"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray30", size = 10)
  )
```